---
title: "Dealing with images in R"
format: gfm
knitr:
  opts_chunk:
    fig.path: "img/images-"
execute:
  echo: true     
  warning: false  
  error: false    
editor_options: 
  chunk_output_type: console
---


# Downloading and handling images

As a first step to image analysis, you need to acquire the source material.
For now, let's assume we have a data frame with image URLs and other relevant information,
for example politicians and gender:


```{r download}
library(tidyverse)

images <- tribble(
  ~name, ~gender, ~url,
    "merkel", "female", "https://upload.wikimedia.org/wikipedia/commons/d/d7/Angela_Merkel_11.jpg",
    "obama", "male", "https://upload.wikimedia.org/wikipedia/commons/d/d1/Barack_Obama_Fold.jpg",
    "rutte", "male", "https://upload.wikimedia.org/wikipedia/commons/c/cd/Mark_Rutte.jpg",  
    "meloni", "female", "https://upload.wikimedia.org/wikipedia/commons/7/78/Giorgia_meloni.jpg",
    "klaebo", "male", "https://upload.wikimedia.org/wikipedia/commons/thumb/1/18/Johannes_H%C3%B8sflot_Kl%C3%A6bo_0736.jpg/500px-Johannes_H%C3%B8sflot_Kl%C3%A6bo_0736.jpg", 
    "vitozzi", "female", "https://upload.wikimedia.org/wikipedia/commons/0/08/Vittozzi_L._%E2%80%93_Biathlon_2023_Nove_Mesto_8374.jpg",
    "schouten", "female", "https://upload.wikimedia.org/wikipedia/commons/0/0d/Irene_schouten-1457901652.jpeg",
    "krol", "male", "https://upload.wikimedia.org/wikipedia/commons/thumb/3/38/2016_World_Single_Distance_Speed_Skating_Championships_-_1500m_M_-_Thomas_Krol.jpg/656px-2016_World_Single_Distance_Speed_Skating_Championships_-_1500m_M_-_Thomas_Krol.jpg"
  )
```

To store the images locally, we can use the `download.file` function. 
To make it easy to run this for each politician, let's define a helper function to download the images as needed:

```{r dlfunction}
download_image <- function(name, url, ...) {
  destfile = paste0("tutorial_images/", name, ".jpg")
  if (!file.exists(destfile)) 
    download.file(url, destfile, mode = "wb", headers=c("User-Agent" = "RTutorial/1.0"))
} 
```

Now, we can use the `pwalk` function to call this function for each row of our data frame.
Note that this is why I included the ellipsis (`...`) in the function arguments:
R will call it with each column from the data frame, and without the ellipsis it would throw an error about the unexpected 'gender' argument.

```{r images}
dir.create("tutorial_images", showWarnings = F)
pwalk(images, download_image, .progress = TRUE)
```

Note that `pwalk` and `pmap` are both ways to call a function on each row of a tibble.
In this case, we use `pwalk` because we are performing an **action** where we care about the 'side effect', 
like downloading a file or printing a plot. `pwalk` returns the *input data*, which makes it perfect for the middle of a pipe since we can just continue with the original data. 

In most cases, we want to use a function to **transform data** rather than execute an action. For this we can use `pmap` (or `map` on single columns/vectors rather than whole data frames). This returns the output of each function call, which is ideal to continue processing the data. 

## Preprocessing images with Magick

Imagemagick is a powerful tool to deal with images which can be accessed with the R `magick` library.

For example, we can use `image_read` and `image_info` to see basic information of our images

```{r imginfo}
library(magick)
paste0("tutorial_images/", images$name, ".jpg") |>
  image_read() |>
  image_info() |> 
  add_column(name=images$name)
```

As you can see, the image sizes vary from ~1 to ~10 megapixels and from portrait to square and even landscape images.
Let's resize all images to be a square 300x300:

```{r resize}
preprocess_image <- function(name, ...) {
  img <- paste0("tutorial_images/", name, ".jpg") |>
    image_read() |>
    image_resize("300x300^") |>
    image_extent("300x300", gravity = "Center", color = "white")
  tibble(name=name, img=list(img))
}
processed_images <- pmap(images, preprocess_image) |> list_rbind() 
images <- left_join(images, processed_images)
images
```

Note the use of `pmap` as before, but now we return a tibble with the name as a column, and the image, wrapped in a list, as a second column.
This allows us the join the processed images with the metadata so we now have all information in one place

Also note this results in **cropped** images because the `^` in `300x300^` instructs Magick to resize the image based on the smallest dimension, ensuring the entire 300x300 area is filled even if the other dimension overflows. The `image_extent` makes the actual change to the canvas.

To see the result, we can simply select a single image:

```{r showimage}
images |> filter(name == "obama") |> pull(img)
```

What happens if you remove the `^` instead?

## Displaying images with ggplot

We can also incorporate them into ggplot with `image_ggplot`,
but this expects only a single image. 
So, let's again make a function to plot one image with the name as title:

```{r imgggplot}
ggplot_politician <- function(name, img, ...) {
  image_ggplot(img)  + 
    labs(title=name |> str_to_title()) + 
    theme(plot.title = element_text(hjust = 0.5))
}
```

Now, we can use `pmap` as before and combine the four plots in a `wrap_plots` call:

```{r patchwork}
library(patchwork)
pmap(images, ggplot_politician) |>
  wrap_plots()
```

# Help me, Chat! Analyzing images with tidyllm

The easiest and perhaps currently most powerful way to do image analysis is with LLMs such as OpenAI's chatGPT
or a local (vision-capable) model which you can run through ollama. In both cases, you can use the tidyllm package to do the actual analysis. 

The example below uses `moondream`, which is a very small vision-capable LLM which you can install through ollama.
To run this example, go to [https://ollama.com/download](https://ollama.com/download) to download ollama,
and then use `ollama pull moondream` to load this model. 

This example uses the same `function - pmap - list_rbind()` pattern used before, which calls the function for each image and returns the result as a data frame that can be joined back on the `name` column in needed.

Note that unfortunately tidyllm cannot use the image directly from our data frame, so we write it to a temporary file which we then delete.
We could of course also write all the preprocessed images to disk rather than keep them in memory. 

```{r describe}
library(tidyllm)

describe_politician <- function(name, img, ...) {
  tmp <- tempfile(fileext = ".png")
  image_write(img, path = tmp, format = "png")
  description <- llm_message(
      "Describe the person in this image",
      .imagefile = tmp
    ) |>
      chat(ollama(.model = "moondream")) |>
      get_reply() 
  unlink(tmp)
  tibble(name=name, description=description)
}

pmap(images, describe_politician) |> list_rbind()
```

We can also run this example on a remote LLM like OpenAI's chatGPT. For this, you would need to 
get an API key from OpenAI and paste it in the line below:

```{r, eval=FALSE}
Sys.setenv(OPENAI_API_KEY="sk-proj-...")
```

After this, we simply replace the `ollama(..)` function with an `openai(..)` function and keep the rest the same.
Interestingly, I did have to change the prompt to avoid triggering a guard rail preventing the AI from commenting on the politician in the picture:

```{r describechat}
#| cache: true
#| dependson: "images"
describe_politician_chat <- function(name, img, ...) {
  tmp <- tempfile(fileext = ".png")
  image_write(img, path = tmp, format = "png")
  description <- llm_message(
      "Describe the main person's clothing, facial expression, and posture in detail. Do not mention their name",
      .imagefile = tmp
    ) |>
      chat(openai(.model = "gpt-4o-mini")) |>
      get_reply() 
  unlink(tmp)
  tibble(name=name, description=description)
}

pmap(images, describe_politician_chat) |> list_rbind()
```

## Using structured output

The function above returned the output of the LLM as a simple text description.
Often, however, we want specific structured information such as the gender, emotion, or setting of a picture.

For this, it is generally best to use **structured outputs**, where we first define the expected output and then tell the AI to provide the requested format. 

First, we define a `schema`, for example with a short description and then the gender, emotion, and apparent age of the person

```{r schema}
face_schema <- tidyllm_schema(
  name = "Face schema",
  description = field_chr(.description = "Short description (1–2 sentences) of the main person shown in the picture"),
  gender = field_fct(.description = "What is the apparent gender of the main person (M for male, F for female)", .levels = c("F", "M")),
  emotion = field_chr(.description = "Single word for the main emotion shown by the person, e.g. happy, sad or angry")
)

```

Now, we can rewrite the function above to include the `.json_schema` (and include the desired output in the prompt).
We also replace `get_reply` by `get_reply_data`, which returns a named list that we can easily turn into a tibble to return:

```{r}
#| cache: true
#| dependson: ["images", "schema"]
analyse_picture <- function(name, img, ...) {
  tmp <- tempfile(fileext = ".png")
  image_write(img, path = tmp, format = "png")
  llm_message(
      "Describe the main person's apparent gender and main emotion as shown in the picture. First give a small description of the overall face, and then list the gender (as F or M) and main emotion",
      .imagefile = tmp
    ) |>
      chat(ollama(.model = "moondream"), .json_schema = face_schema) |>
      get_reply_data() |>
    as_tibble() |>
    add_column(name=name)
}

pmap(images, analyse_picture) |> list_rbind()
```

If you play around with this you will find out that moondream often makes seemingly stupid mistakes. You might get better results with a more powerful model (such as `llava`, `qwen3-vl:8b` or `llama3.2-vision:11b`) but note that it depends on your computer whether you will be able to run them: the bigger models require a lot of GPU memory, which not all computers will support. See https://ollama.com/library for an overview of existing models. 

You can also use proprietary models such as `chatgpt` as used above, but these might hit the brake more often where it comes to talking about existing persons, so you might have to massage the prompt to ensure the AI of your good intentions as a researcher...

## Embedding images

We've discussed using word and document embeddings in our [text embeddings tutorial](embeddings.md).
It is also quite easy to embed images, which allows us to cluster images similar to how a topic model clusters words and documents.

While the contextual text embeddings were performed by calling python behind the scenes with reticulate,
let's now use the R packages `torch` and `torchvision`, which are built directly in R*

(*) Well, actually, the R packages call C++ libraries to do most of the work -- but python's torch package also just calls the same C++ library, so in that sense the R package is more 'native' than the packages that use reticulate to call python. And in fact many of the packages we use daily in R (and python!) actually call libraries written in more efficient languages such as C behind the scene, including many of our beloved tidyverse packages!

```{r}
library(torch)
library(torchvision)
library(magick) # Useful for image loading
```


```{r}
transform <- function(img) {
  img |>
    transform_resize(size = c(224, 224)) |>
    transform_to_tensor() |>
    transform_normalize(mean = c(0.485, 0.456, 0.406), std = c(0.229, 0.224, 0.225))
}
```

```{r}
model <- model_resnet50(pretrained = TRUE)
```

```{r}
model$eval()
autograd_set_grad_mode(FALSE)
```

```{r}
preprocess_image <- function(name) {
  tensor <- str_glue("tutorial_images/{name}.jpg") |>
    image_read() |>
    transform_to_tensor() |>
    transform_resize(size = c(224, 224)) |>
    transform_normalize(
      mean = c(0.485, 0.456, 0.406), 
      std = c(0.229, 0.224, 0.225)
    )
}

embeddings <- map(images$name, preprocess_image) |>
  torch_stack() |>
  model() |>
  as.matrix() |>
  magrittr::set_rownames(images$name)

#' Function to calculate cosine similarity between rows of a matrix
cosine_sim_rows <- function(M) {
  # Step 1: normalize by dividing each row by its Euclidean length
  M_norm <- M / sqrt(rowSums(M^2))
  # Step 2: cosine = inner products of normalized vector pairs = matrix times transposed matrix
  M_norm %*% t(M_norm)
}

similarities <- cosine_sim_rows(embeddings)
ggcorrplot::ggcorrplot(similarities, show.diag=F, lab=T)
```

We can make this look nicer by including the actual images, but that means we need to build the image ourselves rather than relying on ggplot.

Unfortunately, 

```{r}
library(ggimage)
similarities_long <- similarities |>
  as.data.frame() |>
  rownames_to_column("name1") |>
  pivot_longer(-name1, names_to="name2") |>
  filter(name1 != name2) 

diagonal <- images |> 
  select(name1=name, name2=name) |>
  mutate(img_path=str_glue("tutorial_images/{name}.jpg"))

ggplot(similarities_long, aes(x=name1, y=name2)) +
  geom_tile(aes(fill=value)) +
  geom_text(aes(label=round(value,2))) +
  geom_image(data=diagonal, aes(image=img_path)) + 
  scale_fill_gradient(low="white", high="darkgreen") + 
  coord_equal() + 
  theme_minimal()
  
```

```{r}
clustering <- hclust(as.dist(1 - similarities), 
                     method = "average") 

d <- ggdendro::dendro_data(clustering)

clusters <- cutree(clustering, k=2)
clusters <- tibble(label = names(clusters), cluster = as.factor(clusters) )

ggplot() +
  geom_segment(data = d$segments,
               aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_text(data = inner_join(d$labels, clusters),
            aes(x = x, y = y - 0.02, label = label, 
                color=cluster),
            angle = 45, hjust = 1) +
  coord_cartesian(clip = "off") +
  ylab("Cosine distance") + 
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.margin = margin(20, 20, 40, 20),
        legend.position = "none"
  )


```
