---
title: 'R basics: Functions'
author: "Kasper Welbers & Wouter van Atteveldt"
date: "2018-09"
output: 
  github_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(echo = TRUE, results = FALSE, message = FALSE, warning = FALSE)
library(printr)
```

# This tutorial

In the previous tutorial you learned about the basic data types and structures in R. This week, we'll first delve into **Functions**, as the main vehicle for performing operations in R. You'll learn how to use functions, and how to get more functions by installing and using packages. As you will see, all functions are in fact created equal (i.e. they work in the same way), so by learning how to use them, you immediately gain access to countless new techniques.

## Exercise

This document contains a tutorial, and there is an excercise that you can find and complete in the separate **R_basics_3_functions_practise.Rmd** file. You need to open this file in RStudio. 

If after this introduction you want to learn more about *functions*, and in particular want to learn how to build your own functions (which opens up a whole world of possibilities), we recommend [this blog post](https://www.datacamp.com/community/tutorials/functions-in-r-a-tutorial) from DataCamp. 

# Functions in R

## What are functions

There are many correct and formal ways to define what functions are, but for the sake of simplicity we will ignore formalities and instead focus on an informal description of how you can think of functions in R:

* A function has the form: `output = function_name(argument1, argument2, ...)`
    + **function_name** is a name to indicate which function you want to use. It is followed by parentheses.
    + **arguments** are the input of the function, and are inserted within the parentheses. Arguments can be any R object, such as numbers, strings, vectors and data.frames. Multiple arguments can be given, separated by commas. 
    + **output** is anything that is returned by the function, such as vectors, data.frames or the results of a statistical analysis. Some functions do not have output, but produce a visualization or write data to disk.
* The purpose of a function is to make it easy to perform a (large) set of (complex) operations. This is crucial, because
    + It makes code easier to understand. You don't need to see the operations, just the name of the function that performs them
    + You don't need to understand the operations, just how to use the function
 
For example, say that you need to calculate the square root of a number. This is a very common thing to do in statistical analysis, but it actually requires a quite complicated set of operations to perform. This is when you want to use a function, in this case the `sqrt` (square root) function. 

```{r}
sqrt(5)
```
In this example, the function name is `sqrt`. The input is the single argument `5`. If you run this code, it produces the output `2.236068`. Currently, R will simply print this output in your Console, but as you learned before, we can assign this output to a name.

```{r}
square_root = sqrt(5)
```

This simple process of input -> function -> output is essentially how you work with R most of the times. You have data in some form. You provide this data as input to a function, and R generates output. You can assign the output to a name to use it in the next steps, or sometimes the output is a visualization or statistical model that you want to interpret.  

## Using functions

Above you saw the simple function `sqrt()`, that given a single number as input returned a single number as output. As you have also seen in the first week, functions can have multiple arguments as input. Recall the following function from the `quanteda` package. You don't have to run the code this time, just try to recognize the arguments. 

```{r, eval=F}
dfm(x = "some text", tolower = TRUE, stem=TRUE)
```

This function, with the name `dfm`, is given several arguments here: `x`, `tolower` and `stem`. Given this input, many operations are performed behind the scenes to create a document-term matrix.

By now we hope you have realized just how broad the use of functions is. The *R syntax* for performing basic mathematical operations such as `sqrt()` is essentially the same as the syntax for creating a document-term matrix, performing advances statistical analysis or creating powerfull visualizations. Accordingly, if you understand this syntax, you can do almost anything in R. 

The many R packages that you can install are mostly just collections of functions (some also provide new **classes**, which we'll save for later). We will now show how you can learn how to use each function by knowing how to view and interpret it's documentation page.

### Viewing and interpreting function documentation

You can access the documentation of a function by typing a question mark in front of the function name, and running the line of code. Let's do this to view the documentation of the `sqrt()` function

```{r}
?sqrt
```

If you run this in RStudio, the help page will pop-up in the bottom-right corner, under the *Help* tab page. Sometimes, if the name of a documentation page is used in multiple packages, you will first receive a list of these packages from which you will have to select the page. 

For the `sqrt()` function, the help page has the **title** "Miscellaneous Mathematical Functions".
Just below the title, you see the **Description**, in which the author of a function briefly describes what the function is for. 
Here we see that there are two functions that are grouped under "Miscellaneous Mathematical Functions", the `abs()` function for computing the absolute value of a number `x`, and the `sqrt()` function for the square root.  

Under description, the **Usage** is shown. This is simply the name of the function or functions, and the possible arguments that you can use. Here the Usage is extremely simple: both functions only take one argument named `x`. In a minute, we'll discuss functions with multiple arguments.

Below usage, the **Arguments** section explains how to use each argument. Here, the only argument is `x`, and it is explained that x is "a numeric or complex vector or array". For now, let's focus only on the case of a numeric vector. It appears that in addition to giving a single value like above (recall that in R this is actually a vector of length 1) we can give a vector with multiple numbers.

```{r}
sqrt(c(1,2,3,4,5))
```

There are more parts to the documentation that we'll ignore for now. Notable parts to look into for yourself are **Details**, that provides more information, and the **Examples** section at the very bottom, which is a great starting point to see a function in action. 

### Understanding multi-argument functions

Now, let's move to a function with multiple arguments. We'll look at the `dfm()` function from the `quanteda` package. To access this function, we first run `library(quanteda)`, to tell R that we want to be able to access the functions in this package. Note that you have to have the package installed as well. This should still be the case from a prior tutorial, but if you changed computers, you will have to run the line `install.packages('quanteda')` first.  

```{r}
library(quanteda)
?dfm
```

First note that the title and description nicely summarize what this function is for: creating a document-feature matrix. Now, when we look at the **Usage** section, we see that there are multiple arguments given between the parentheses, and all these arguments are explained in the **Arguments** section.

An important part of the usage syntax, that we haven't seen in the `sqrt()` function, is that all arguments other than `x` have a value assigned to them, in the form `argument = value`. The argument `tolower` has the value `TRUE`, `stem` has the value `FALSE`, etc. 

These are the default values for these argument, that are used if the user does not specify them. This way, We can use the `dfm()` function with the default settings by only entering the `x` argument.

```{r}
example_texts = c("Some example text", "Some more text")
dfm(example_texts)
```

If we run this line of code, it returns a matrix with the frequencies of each word for each text. Note that the word "Some" in both texts has been made lowercase, because the `tolower` argument (that is described as "convert all features to lowercase") is `TRUE` by default. 

Arguments that don't have a default value, such as `x` in the `dfm()` function, are mandatory. Running the following line of code will give the error `argument "x" is missing, with no default`.

```{r, eval=F}
dfm()
```

It is often the case that in addition to the mandatory arguments you want to specify some specific other arguments. 
For this, there are two ways to *pass* arguments to a function.
 
* Use the same order in which they are specified in **Usage**
* Pass the arguments with their respective names

To demonstrate passing by order, let's run the `dfm()` function again, but this time with input for `tolower` and `stem`. 

```{r}
dfm(example_texts, TRUE, TRUE)
```

In the output we see that the word "example" has been `stemmed` to "examp", because we have set the `stem` argument to `TRUE`. The words are still made lowercase, since we passed `TRUE` to `tolower`, which was also the default value. 

Passing by order is annoying if you want to specify only one particular argument. In the current example, we had to explicitly pass TRUE to `tolower` even though this was already the default. More importantly, this can become confusing and cause mistakes if you pass many arguments. Therefore, it is often recommended to pass values by name. Here we use this to only change `stem` to `TRUE`.

```{r}
dfm(x = example_texts, stem = TRUE)
```

Overall, passing by name is more explicit and safe, but it can be needlessly verbose to specify all names, such as `x = example_texts` in the example. Thus, we can combine both approaches, passing the arguments to the left (i.e. the first, and often mandatory, arguments) by order, and arguments further to the right by name. 

```{r}
dfm(example_texts, stem=TRUE)
```

Whatever approach you prefer, try to be consistent, and take into account whether your code will still be easy to interpret for other that you share it with, or for yourself in the future. A good general rule is to pass mandatory arguments (such as `x` in the `dfm` function) without a name, but all the optional arguments (that have a default, such as `tolower` and `stem`) by name.

### A small note about 'methods' and 'generic functions'

Some functions are *generic functions*, that use different *methods* depending on the input that they are used with. Ignoring technicalities, there's one thing you currently need to know about them, because you will need it to interpret their documentation pages. 

A method is a function that is associated with a specific object. For example, subsetting a `vector` works differently from subsetting a `data.frame` or `matrix`. Still, it is convenient to only have one function called `subset()` that can be used on all these kinds of input. In R, the `subset()` functions is therefore a *generic function*, that will behave differently depending on the kind of input. 

The type of input to the `subset()` function therefore affects what type of arguments can be used. You can see this in the documentation page.

```{r}
?subset
```

In the description we see that `subset()` can be used on vectors, matrices or data.frames. The **Usage** section therefore contains different versions, for different *S3 methods* (ignore "S3" for now) that are associated with different kinds of input. The general form is `subset(x, ...)`, which shows that subset always requires an argument `x`, and in the **Arguments** we see that `x` is the "object to be subsetted". We then see three methods: default, 'matrix' and 'data.frame'.

* The default will be used if `x` is neither a `matrix` or `data.frame` (for instance a `vector`). In this case the only argument is *subset*, which is the expression (e.g., `x > 10`) used to make a selection. 
* If the input is a 'matrix', there are two additional arguments: *select* and *drop*. It makes sense that these are not available for vectors, because they are both only relevant if there are multiple columns. That is, *select* is used for selecting columns, and *drop* can be used to have subset return a `vector` (instead of a `matrix`) if only one row or column remains after subsetting.
* If the input is a 'data.frame', the same arguments are used as for 'matrix' (but internally the method works differently)

### The special case of the three dot ellipsis

A special type of argument that you'll often encounter in function documentation is the three dot ellipsis (`...`). This is used to pass any number of named or unnamed arguments. A good example of how this is used, is in the `data.frame()` function. In a previous tutorial you saw that you can use this function to create a data.frame from vectors, where names are used as column names. Now, you will see that these are actually just *named arguments*.

```{r}
?data.frame
data.frame(x = 1:5, y = c('a','b','c','d','e'))
```

As an additional example, consider the `sum()` function. 
Here the `...` is used for "numeric or complex or logical vector". This means that we can add any number of arguments with numbers in them, and they will all be added up.

```{r}
?sum
sum(1, 2, 3, c(1,2,3))
```

To clarify, if we want to set any of the other arguments, such as `na.rm` in `sum()`, we can do so by referring to them by name. By default, `sum()` returns NA (R's way of saying "missing") if any NA is present. As noted in the documentation, we can ignore the NA values by setting `na.rm` to `TRUE`

```{r}
sum(1,2,NA)
sum(1,2,NA, na.rm = T)
?dfm
```

Finally, a way in which the three dot ellipsis is also often used, is to pass arguments on to another function that is used within the function. If you look back at the documentation for the `dfm()` function, you'll see in the explanation of `...`: "additional arguments passed to tokens; not used when x is a dfm". In this case, you can see which arguments these are by looking at the documentation of the `tokens` function. Here you see that you could also pass the argument `remove_numbers = TRUE` to `dfm()`.


