---
title: "Supervised Text Classification"
author: "Wouter van Atteveldt & Kasper Welbers"
date: "2022-01"
output: 
  github_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r, echo=F, message=F}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, results=F, fig.keep='none')
library(printr)
```

This handout contains a brief introduction to using supervised machine learning for text classification in R. 
In particular, we will show how both [quanteda.textmodels](https://github.com/quanteda/quanteda.textmodels) and [tidymodels](machine_learning.md) can be used
to train and fit supervised text classification models. 

In this tutorial, we use the following packages:
(note that there is no need to install packages again if you previously installed them)

To install the various packages (of course, you can skip this step for packages already installed on your machine):

```{r, eval=F}
install.packages("tidyverse")
install.packages("tidymodels")
install.packages("textrecipes")
install.packages("quanteda")
install.packages("quanteda.textmodels")
```


## Getting training data

For machine learning, we need annotated training data. Fortunately, there are many review data files available for free.
For this exercise, we will use a set of Amazon movie reviews cached as CSV on our github site. 
See http://deepyeti.ucsd.edu/jianmo/amazon/index.html for other (and more up-to-date) Amazon product reviews.

```{r}
library(tidyverse)
reviews = read_csv("https://raw.githubusercontent.com/ccs-amsterdam/r-course-material/master/data/reviews.csv")
head(reviews)
table(reviews$overall)
```

As you can see, there's the star rating (`overall`), a summary text, the full review text, and the ID of the reviewer and the product (ASIN). 

Before getting started, let's define a two-class rating from the numeric overall rating:

```{r}
reviews = mutate(reviews, rating=as.factor(ifelse(overall==5, "good", "bad")))
```

The goal for this tutorial is supervised sentiment analysis, i.e. to predict the star rating given the review text. 

# Supervised text analysis with quanteda

To get started, let's use quanteda to do the textual preprocessing and `quanteda.textmodels` to run the machine learning. 

First, let's create a corpus using both summary and reviewtext, and creating a two-way class `rating` in addition to the numeric `overall` score:

```{r}
library(quanteda)
review_corpus = reviews |> 
  mutate(text = paste(summary, reviewText, sep="\n\n"),
         doc_id = paste(asin, reviewerID, sep=":")) |>
  select(-summary, -reviewText) |>
  corpus()
```

Now, we can split the corpus into test and train set: (using `set.seed` for reproducibility)

```{r}
set.seed(1)
testset = sample(docnames(review_corpus), 2000)
reviews_test =  corpus_subset(review_corpus, docnames(review_corpus) %in% testset)
reviews_train = corpus_subset(review_corpus, !docnames(review_corpus) %in% testset)
```

## Training the model using quanteda

First, we create a dfm from the training data, using stemming and trimming:

```{r}
dfm_train = reviews_train |> 
  tokens() |>
  dfm() |> 
  dfm_wordstem(language='english') |>
  dfm_select(min_nchar = 2) |>
  dfm_trim(min_docfreq=10)
```

And now we can train a model, for example a naive bayes model. 

```{r}
library(quanteda.textmodels)
rating_train = docvars(reviews_train, field="rating")
m_nb <- textmodel_nb(dfm_train, rating_train)
```

We can also inspect the NB parameters directly to find out which words are the best predictors.
The parameters in the model are the 'class conditional posterior estimates', or `P(w|C)`.
We are interested in the opposite, but can use Bayes' theorem `P(C|w)=P(w|C)P(C)/P(w)`.
We also know that `P(w)=Sum(P(w|C)P(C))`, so if we assume an equal prior distribution of the classes we can simplify to
`P(good|w)=P(w|good)/(P(w|good) + P(w|bad))`:

```{r}
scores = t(m_nb$param) |> 
  as_tibble(rownames = "word") |> 
  mutate(relfreq=bad+good,
         bad=bad/relfreq,
         good=good/relfreq) 
scores |>
  filter(relfreq > .0001) |>
  arrange(-bad) |>
  head()
```

So, negative predictors are flare, energ(y), and useless. Positive predictors include awesom(e), amaz(ing), and excel(lent). 

## Testing the model

To see how well the model does, we test it on the test data.

For this, it's important that the test data uses the same features (vocabulary) as the training data
The model contains parameters for these features, not for words that only occur in the test data.
So, we also skip the selection and trimming steps, and instead adapt the test vocabulary (dfm columns)
to those in the train set:

```{r}
dfm_test = reviews_test |> 
  tokens() |>
  dfm() |> 
  dfm_wordstem(language='english') |>
  dfm_match(featnames(dfm_train))
```

Now, we can predict the classes of the test data:

```{r}
nb_pred <- predict(m_nb, newdata = dfm_test)
head(nb_pred)
```

To see how well we do, we can compare the predicted sentiment to the actual sentiment:

```{r}
predictions = docvars(reviews_test) |>
  as_tibble() |>
  add_column(prediction=nb_pred)
predictions |> 
  mutate(correct=prediction == rating) |>
  summarize(accuracy=mean(correct))
```

So (at least on my computer), 75% accuracy. Not bad for a first try -- but movie reviews are quite simple, this will be a lot harder for most political text...

We can also use the yardstick metrics from tidymodels:

```{r}
library(tidymodels)
metrics = metric_set(accuracy, precision, recall, f_meas)
metrics(predictions, truth = rating, estimate = prediction)
conf_mat(predictions, truth = rating, estimate = prediction)
```

# Aside: Scaling with Quanteda

Quanteda also allows for supervised and unsupervised scaling.
Although wordfish is an unsupervised method (so doesn't really belong to this tutorial), it produces a nice visualization:

```{r}
library(quanteda.textplots)
set.seed(1)
m_wf <- textmodel_wordfish(dfm_train, sparse=T)
bestwords = scores |> 
  mutate(value=pmax(good, bad)) |>
  group_by(round(log10(relfreq))) |>
  slice_max(value, n=15)

textplot_scale1d(m_wf, margin = "features", highlighted = bestwords$word)
```

Wordfish automatically scales all documents according to an underlying latent variable (the beta),
which is assumed to correspond to the most interesting dimension in the data. 
To make this easier to read, I highlight the words with the most explanatory value in the naive bayes model,
taking a stratified sample of 15 words per frequency band. 

# Supervised machine learning with tidymodels

Using the preprocessing steps from [textrecipes](https://textrecipes.tidymodels.org),
we can also use tidymodels to test our data.

Although this involves a bit more steps if you are already using quanteda, 
using tidymodels allows more flexibility in selecting and tuning the best models. 

The example below will quickly show how to train and test a model using these recipes.
See the [machine learning with Tidymodels](machine_learning.md) handout and/or the
[tidyverse documentation](https://tidyverse.org) for more information. 

## Using `textrecipes` to turn text into features

```{r}
library(tidymodels)
library(textrecipes)
rec = recipe(rating ~ summary + reviewText, data=reviews) |>
  step_tokenize(all_predictors())  |>
  step_tokenfilter(all_predictors(), min_times = 3) |>
  step_tf(all_predictors())
```

We can inspect the results of the preprocessing by `prepping` the recipe and baking the training data:

```{r}
rec |> 
  prep(reviews) |>
  bake(new_data=NULL) |> 
  select(1:10)
```


## Fitting and testing a model

First, we create a *worflow* from the recipe and model specification. 
Let's start with a naive bayes model:

```{r}
library(discrim)
lr_workflow = workflow() |>
  add_recipe(rec) |>
  add_model(logistic_reg(mixture = 0, penalty = 0.1))
```   

Now, we can split our data, fit the model on the train data, and validate it on the test data:

```{r}
split = initial_split(reviews)
m <- fit(lr_workflow, data = training(split))
predict(m, new_data=testing(split)) |>
  bind_cols(select(testing(split), rating)) |>
  rename(predicted=.pred_class, actual=rating) |>
  metrics(truth = actual, estimate = predicted)
```

So, we get a slightly better accuracy than with the naive bayes model tested above.
To see which words are the most important predictors, we can use the `vip` package
to extract the predictors, and then use regular tidyverse/ggplot functions to visualize it:

```{r}
m |> extract_fit_parsnip() |>
  vip::vi() |> 
  group_by(Sign) |>
  top_n(20, wt = abs(Importance)) %>%
  ungroup() |>
  mutate(
    Importance = abs(Importance),
    Variable = str_remove(Variable, "tf_"),
    Variable = fct_reorder(Variable, Importance)
  ) |>
  ggplot(aes(x = Importance, y = Variable, fill = Sign)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Sign, scales = "free_y") +
  labs(y = NULL)
```

The positive predictors make perfect sense: *great*, *best*, *excellent*, etc. 
So, interestingly *not*, *but*, and *ok* are the best negative predictors, and *good* in the summary is also not a good sign. 
This makes it interesting to see if using ngrams will help performance, as it is quite possible that it is *not good*, rather than good. 
Have a look at the [textrecipes documentation](https://textrecipes.tidymodels.org/reference/) to see the possibilities for text preprocessing.

Also, we just tried out a regularization penalty of 0.1, and it is quite possible that this is not the best choice possible.
Thus, it is a good idea to now do some hyperparameter tuning for the regularization penalty and other parameters. 
Take a look at the [machine learning handout](machine_learning.md) and/or the [tune documentation](https://tune.tidymodels.org/)
to see how to do parameter tuning. 

Of course, you can also try one of the other classification models in [parsnip](https://parsnip.tidymodels.org/),
and/or try a regression model instead to predict the actual star value. 