---
title: "Fitting LDA Models in R"
author: "Wouter van Atteveldt & Kasper Welbers"
date: "2020-03"
output: 
  github_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r, echo=F, message=F, warning=F}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.path = "img/")
library(printr)
```

# Introduction

LDA, which stands for Latent Dirichlet Allocation, is one of the most popular approaches for probabilistic topic modeling. The goal of topic modeling is to automatically assign topics to documents without requiring human supervision. Although the idea of an algorithm figuring out topics might sound close to magical (mostly because people have too high expectations of what these 'topics' are), and the mathematics might be a bit challenging, it is actually really simple fit an LDA topic model in R.  

A good first step towards understanding what topic models are and how they can be useful, is to simply play around with them, so that's what we'll do here. First, let's create a document term matrix from the inaugural speeches in quanteda, at the paragraph level since we can expect these to be mostly about the same topic:

```{r}
library(quanteda)
corp = corpus_reshape(data_corpus_inaugural, to = "paragraphs")
dfm = dfm(corp, remove_punct=T, remove=stopwords("english"))
dfm = dfm_trim(dfm, min_docfreq = 5)
```

To run LDA from a dfm, first convert to the topicmodels format, and then run LDA. 
Note the useof `set.seed(.)` to make sure that the analysis is reproducible. 

```{r}
library(topicmodels)
dtm = convert(dfm, to = "topicmodels") 
set.seed(1)
m = LDA(dtm, method = "Gibbs", k = 10,  control = list(alpha = 0.1))
m
```

Although LDA will figure out the topics, we do need to decide ourselves how many topics we want. Also, there are certain hyperparameters (alpha) that we can tinker with to have some control over the topic distributions. For now, we won't go into details, but do note that we could also have asked for 100 topics, and our results would have been much different.  

## Inspecting LDA results

We can use `terms` to look at the top terms per topic:

```{r}
terms(m, 5)
```

The `posterior` function gives the posterior distribution of words and documents to topics,
which can be used to plot a word cloud of terms proportional to their occurrence:

```{r}
topic = 6
words = posterior(m)$terms[topic, ]
topwords = head(sort(words, decreasing = T), n=50)
head(topwords)
```

Now we can plot these words:

```{r lda-wordcloud}
library(wordcloud)
wordcloud(names(topwords), topwords)
```

We can also look at the topics per document, to find the top documents per topic:

```{r}
topic.docs = posterior(m)$topics[, topic] 
topic.docs = sort(topic.docs, decreasing=T)
head(topic.docs)
```

Given the document ids of the top documents, we can look up the text in the `corp` corpus

```{r}
topdoc = names(topic.docs)[1]
topdoc_corp = corp[docnames(corp) == topdoc]
texts(topdoc_corp)
```

Finally, we can see which president preferred which topics:

```{r lda-heatmap}
docs = docvars(dfm)[match(rownames(dtm), docnames(dfm)),]
tpp = aggregate(posterior(m)$topics, by=docs["President"], mean)
rownames(tpp) = tpp$President
heatmap(as.matrix(tpp[-1]))
```

As you can see, the topics form a sort of 'block' distribution, with more modern presidents and older presidents
using quite different topics. So, either the role of presidents changed, or language use changed, or (probably) both.

To get a better fit of such temporal dynamics, see the session on *structural topic models*, which allow you to condition topic proportions and/or contents on metadata covariates such as source or date. 

## Visualizing LDA with LDAvis

`LDAvis` is a nice interactive visualization of LDA results.
It needs the LDA and DTM information in a slightly different format than what's readily available, but you can use the code below to create that format from the
lda model `m` and the `dtm`.
If you don't have it yet, you'll have to install the `LDAvis` package, and you might also have to install the `servr` package.

```{r, eval=F}
library(LDAvis)   

dtm = dtm[slam::row_sums(dtm) > 0, ]
phi = as.matrix(posterior(m)$terms)
theta <- as.matrix(posterior(m)$topics)
vocab <- colnames(phi)
doc.length = slam::row_sums(dtm)
term.freq = slam::col_sums(dtm)[match(vocab, colnames(dtm))]

json = createJSON(phi = phi, theta = theta, vocab = vocab,
     doc.length = doc.length, term.frequency = term.freq)
serVis(json)
```
