---
title: "Simple Statistical Modeling in R"
author: "Wouter van Atteveldt & Kasper Welbers"
date: "November 2018"
output: 
  github_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---



```{r, echo=F, message=F}
knitr::opts_chunk$set(eval=T, warning=FALSE, message=FALSE, fig.path = "img/")
```

Basic Modeling
===

In this tutorial we use a file adapted from the data published by Thomas Piketty as a digital appendix to his book "Capital in the 21st Century". You can find the original files here: http://piketty.pse.ens.fr/files/capital21c/en/xls/, but to make things easier we've published a cleaned version of this data set on our repository. 


```{r}
library(tidyverse)
url = "https://raw.githubusercontent.com/ccs-amsterdam/r-course-material/master/data/piketty_capital.csv"
capital = read_csv(url)
head(capital)
```

This data set describes the accumulation of public and private capital per year for a number of countries, expressed as percentage of GDP. So, in Australia in 1970, the net assets owned by the state amounted to 61% of GDP. 

In this tutorial we mainly use the `stats` package. This is loaded by default, so you do not need to call `library(stats)`.

T-tests
===

First, let's split our countries into two groups, anglo-saxon countries and european countries (plus Japan):
We can use the `ifelse` command here combined with the `%in%` operator

```{r}
anglo = c("U.S.", "U.K.", "Canada", "Australia")
capital = mutate(capital, Group = ifelse(capital$Country %in% anglo, "anglo", "european"))
table(capital$Group)
```

Now, let's see whether capital accumulation is different between these two groups. 
We use an (independent samples) T-test, where we use the *formula notation* (`dependent ~ independent`) 
to describe the model we try to test. 

```{r}
t.test(capital$Private ~ capital$Group)
```

So, according to this test capital accumulation is indeed significantly higher in European countries than in Anglo-Saxon countries.

Of course, the data here are not independently distributed since the data in the same year in different countries is related
(as are data in subsequent years in the same country, but let's ignore that for the moment)
We could also do a paired t-test of average accumulation per year per group by first using the cast command to aggregate the data.
Note that we first remove the NA values (for Spain).

```{r}
pergroup = capital %>% 
  na.omit() %>% 
  group_by(Year, Group) %>% 
  summarize(Private=mean(Private))
```

Let's plot the data to have a look at the lines:

```{r modeling_plot}
library(ggplot2)
pergroup %>% 
  ggplot + geom_line(aes(x=Year, y=Private, colour=Group))
```

So initially capital is higher in the Anglo-Saxon countries, but the European countries overtake quickly and stay higher.

Now, we can do a paired-sample t-test. This requires the group measurements to be in the same row across different columns, so that the 'anglo' and 'european' are seen as two 'measurements' on the same year. We therefore first use pivot_wider, as discussed in the tutorial on reshaping data:

```{r}
pergroup = pivot_wider(pergroup, names_from = Group, values_from = Private)
```

Now we can do a t.test of two different columns, using the `data$column` notation to specify columns:

```{r}
t.test(pergroup$anglo, pergroup$european, paired=T)
```

So, the mean difference per year between the groups is indeed significant; t(40) = -6.533, p < 0.001. 

Finally, when reporting a t-test, you will want to report the means and standard deviations. 
Since we have the vectors for anglo and european, we can simply use the `mean` and `sd` functions.
With the argument `na.rm = T` (NA remove is TRUE) we say that missing values are ignored (otherwise, the `mean` and `sd` functions would return NA if the vectors have any NA values).

```{r}
mean(pergroup$anglo, na.rm = T)
sd(pergroup$anglo, na.rm = T)

mean(pergroup$european, na.rm = T)
sd(pergroup$european, na.rm = T)
```

So now we can report that the private capital accumulation per year was lower for Anglo-Saxon countries (M=3.75, SD=0.61) compared to European countries (M=4.21; SD=1.00), and this difference is significant; t(40) = -6.533, p < 0.001. 

Anova
===

We can also use a one-way Anova to see whether accumulation differs per country.
Let's first do a box-plot to see how different the countries are.

Base-R `plot` by default gives a box plot of a formula with a nominal independent variable.
For this, we first need to tell R that Country is a factor (nomimal) rather than textual variable

```{r modeling_anova}
capital = mutate(capital, Country = as.factor(Country))
plot(capital$Private ~ capital$Country)
```

So, it seems that in fact a lot of countries are quite similar, with some extreme cases of high capital accumulation.
(also, it seems that including Japan in the European countries might have been a mistake).
Note that if you are not seeing all country names on the x-axis, you should make the plotting window wider (i.e. drag the window borders). 

We use the `aov` function for this. There is also a function named `anova`, but this is meant to analyze already fitted models, 
as will be shown below.

```{r}
m = aov(capital$Private ~ capital$Country)
summary(m)
```

So in fact there is a significant difference. 
However, the Anova only tells us that there is a difference between groups.
From the box plot we get some idea of the differences between specific countries, but often we also want to know whether there is a significant difference between specific countries.
For this we can use the `pairwise.t.test` to perform post-hoc comparisons:

```{r}
posthoc = pairwise.t.test(capital$Private, capital$Country, p.adj = "bonf")
round(posthoc$p.value, 3)
```

When you report the results of an Anova, you will often want to report specific means and standard deviations. 
This is simply a matter of aggregating (i.e. summarizing) the Private capital accumulation grouped by countries.

```{r}
capital %>%
  group_by(Country) %>%
  summarise(M = mean(Private, na.rm = T), SD = sd(Private, na.rm = T))
```


Linear models (linear regression analysis)
===

A more generic way of fitting models is using the `lm` command.
This is how you do ordinary linear regression analysis in R (but more generally, `aov` is also a linear model).
Let's see how well we can predict the `capital` variable (dependent) by the `country` and `public capital` variables (independent).

The lm function also takes a formula as the first argument. 
The format is `dependent ~ independent1 + independent2 + ...`.

```{r}
m = lm(Private ~ Country + Public, data=capital)  
summary(m)
```

As you can see, R automatically creates dummy values for the nominal variable `Country`, using the first value (`U.S.`) as reference category.
An alternative is to remove the intercept and create a dummy for each country:

```{r}
m = lm(Private ~ -1 + Country + Public, data=capital)
summary(m)
```

(`- 1` removes the intercept because there is an implicit +1 constant for the intercept in the regression formula)

You can also introduce interaction terms by using either the `:` operator (which only creates the interaction term)
or the `*` (which creates a full model including the main effects).
To keep the model somewhat parsimonious, let's use the country group rather than the country itself

```{r}
m1 = lm(Private ~ Group + Public, data=capital)
m2 = lm(Private ~ Group + Public + Group:Public, data=capital)
```

Here we have created two models.
A common way to investigate (and report) these models is by showing them side by side in a table.
There are several packages in R for making nice regression tables (e.g., sjPlot, texreg, stargazer).
Here we'll use the sjPlot package. 

```{r}
## remember to first install with install.packages('sjPlot')
library(sjPlot)
tab_model(m1, m2)
```

So, there is a significant interaction effect which displaces the main effect of public wealth. 

One of the cool things about sjPlot is that the table is produced in HTML, which makes it easy to copy/paste the table into a .DOC file (if you're working with latex, the `texreg` package is more convenient).

```{r}
tab_model(m1,m2, file = 'model.html')
browseURL('model.html')
```


Comparing and diagnosing models
===

A relevant question can be whether a model with an interaction effect is in fact a better model than the model without the interaction.
This can be investigated with an anova of the model fits of the two models:

```{r}
m1 = lm(Private ~ Group + Public, data=capital)
m2 = lm(Private ~ Group + Public + Group:Public, data=capital)
anova(m1, m2)
```

So, the interaction term is in fact a significant improvement of the model. 
Apparently, in European countries private capital is accumulated faster in those times that the government goes into depth. 

After doing a linear model it is a good idea to do some diagnostics.
We can ask R for a set of standard plots by simply calling `plot` on the model fit.
This will actually create 4 plots, so you'd have to hit `Return` to see them all.
A nice alternative is to tell R to split the plotting window into multiple rows and columns.
With the parameter (`par`) `mfrow`, we here set the number of rows and columns to 2, to created a 2x2 grid.

```{r modeling_lmdiag}
par(mfrow=c(2,2))
plot(m)
par(mfrow=c(1,1))
```

Note that after plotting we immediately return the 2x2 grid to a 1x1 grid, so that for the next plot we use the default settings.

If these diagnostic plots are new to you, a good explanation can be found on: https://data.library.virginia.edu/diagnostic-plots/
Also, see http://www.statmethods.net/stats/rdiagnostics.html for a more exhausitve list of model diagnostics.



Jamovi
===

R features several packages with alternative implementations of basic statistics. One of these is the `jmv` package, which allows you to use the stats functions from [Jamovi](https://www.jamovi.org). 
Jamovi is an open-source statistical spreadsheet program that runs on R.
You can also run it as a GUI (think separate program) that is rather similar to SPSS (i.e. you just click stuff).
It can be a useful stepping stone from SPSS to R.

Note that installing jmv might take a while.

```{r eval=F}
install.packages('jmv')
```

The ANOVA function in jmv gives output very similar to the SPSS output.

```{r}
library(jmv)

ANOVA(capital, dep = 'Private', factors = 'Country', postHoc = 'Country')
```

Likewise for t-tests. Here we compute an independent samples (IS) t-test.

```{r}
ttestIS(capital, vars = 'Private', group = 'Group', plots=T)
```

