---
title: "Topic modeling with Tidy Text"
author: "Wouter van Atteveldt"
date: "2022-10"
output: 
  github_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = FALSE, message = FALSE, warning = FALSE, fig.keep='none')
library(printr)
```

# Introduction

LDA, which stands for Latent Dirichlet Allocation, is one of the most popular approaches for probabilistic topic modeling. The goal of topic modeling is to automatically assign topics to documents without requiring human supervision. Although the idea of an algorithm figuring out topics might sound close to magical (mostly because people have too high expectations of what these ‘topics’ are), and the mathematics might be a bit challenging, it is actually really simple fit an LDA topic model in R.

A good first step towards understanding what topic models are and how they can be useful, is to simply play around with them, so that’s what we’ll do here. 
For more information, see also the material on my 2019 [GESIS topic modeling workshop](https://github.com/vanatteveldt/gesis-topic-models).


The process of topic modeling is:

 1. Clean up your corpus and create a DTM or document-term matrix
 2. Fit the topic model
 3. Validate and analyze the resulting model

# (1) Creating a DTM

We use the State of the Union speeches, and we split them per paragraph assuming that these are somewhat consistent in their topic.

Note: For more information on basic tidytext usage, see [our tidytext tutorial](https://github.com/ccs-amsterdam/r-course-material/blob/master/tutorials/tidytext.md) and/or the [official tidytext tutorial](https://www.tidytextmining.com/). 

```{r}
library(sotu)
library(tidyverse)
library(tidytext)
sotu_paragraphs = add_column(sotu_meta, text=sotu_text) |> 
  as_tibble() |>
  filter(year > 1945) |>
  unnest_tokens(text, text, token='paragraphs') |>
  mutate(doc_id=row_number())
```

Next, we create a token list and do some light cleaning:

```{r}
sotu_tokens = sotu_paragraphs |> 
  unnest_tokens(word, text)
sotu_tokens_clean = sotu_tokens |> 
  anti_join(tidytext::stop_words) |>
  group_by(word) |>
  filter(n() > 10,
         !str_detect(word, "[^a-z]")) |>
  ungroup()
print(glue::glue("{length(unique(sotu_tokens_clean$word))} unique words in ",
                 "{length(unique(sotu_tokens_clean$doc_id))} documents"))
```

Now we can create a document-term matrix. This essentially a (sparse) matrix showing
how often each term (column) occurs in each document (row):

```{r}
dtm = sotu_tokens_clean |> 
  group_by(doc_id, word) |>
  summarize(n=n()) |>
  cast_dtm(doc_id, word, n)
```

We can inspect a corner of the dtm by casting it to a regular (dense) matrix:

```{r}
as.matrix(dtm[1:5, 1:5])
```

# (2) Running the topic model

We can now fit the topic model from the dtm using the `LDA` function in the topicmodels package.
Note that we use `set.seed` to create a reproducible setup since topic modeling is deterministic.
Note also that we set a relatively low alpha parameter, see our [understanding alpha](https://github.com/ccs-amsterdam/r-course-material/blob/master/tutorials/understanding_alpha.md) handout for more information.

```{r, cache=TRUE}
library(topicmodels)
set.seed(123)
m = LDA(dtm, method = "Gibbs", k = 10,  control = list(alpha = 0.1))
```

# (3) Inspecting and analysing the results

Now that we have run the model, we can use the `terms` function to check the top terms:

```{r}
terms(m, 5)
```

So, we can see some pretty clear clusters, e.g. about health care and the military. 

## Word-topic probabilities

To extract how strongly each word is associated with each topic, we can use the `tidy` function,
which turns statistical modeling results (e.g. `m`) into tidy tables.

```{r}
# You might have to run install.packages("reshape2")
words = tidy(m)
head(words)
```

This allows us to e.g. plot a word cloud for each topic:

```{R}
library(ggwordcloud)
words |> 
  group_by(topic) |>
  slice_max(beta, n=25) |>
  ggplot() + 
  geom_text_wordcloud(aes(label=term, size=beta)) +
  facet_wrap(vars(topic), nrow=2)
```


We can also compare the words in two topics by computing the log difference in their betas. 
For example, to compare the energy (10) and economy (8) topics and see which word is most typical
in that topic compared to the other topic:

```{r}
words |> filter(topic %in% c(8, 10)) |>
  pivot_wider(names_from=topic, values_from=beta) |>
  filter(`8`>.001 | `10` > .001) |>
  mutate(log_ratio = log2(`8`/`10`)) |>
  slice_max(abs(log_ratio), n=20) |>
  ggplot() + geom_col(aes(x=log_ratio, y=fct_reorder(term, log_ratio)))
```


## Topics per document

Similary to above, we can also extract to topics per document:

```{r}
topics = tidy(m, matrix='gamma')
```

We can join this back with the original metadata
(note that we convert doc_id to character here since DTM row names are always character values):

```{r}
meta = sotu_paragraphs |> 
  mutate(document=as.character(doc_id)) |>
  select(document, year, president, party)
topics = tidy(m, matrix='gamma') |> inner_join(meta)
head(topics)
```

Now, we can e.g. compare topic usage per party:

```{r}
topics |> 
  group_by(party, topic) |> 
  summarize(gamma=mean(gamma)) |> 
  ggplot() + geom_col(aes(x=gamma, y=as.factor(topic), fill=party), position=position_dodge())  +
  scale_fill_manual(values=list(Democratic='#5555ff', Republican='#ff5555'))
```

So, it seems that democrats prefer topic 9 (trade) and 5 (jobs), while republicans prefer topics 6 (law and order) and 2 (american values). 

## Topics in their original  context

Finally, we can use `augment` to extract the topic assignment for each word in each document:

```{r}
assignments = augment(m, data=dtm)
head(assignments)
```

We can join this back with the original data frame to get the meta data per document
(converting the doc_id column to character since row names are always characters,
hence the dtm always has textual document ids)


```{r}
assignments = augment(m, data=dtm) |> 
  mutate(doc_id=as.numeric(document)) |> 
  select(doc_id, word=term, topic=.topic)
assignments = left_join(sotu_tokens, assignments)
head(assignments)
```

This can be used to inspect topic usage within documents using the [tokenbrowser](https://github.com/kasperwelbers/tokenbrowser) package developed by Kasper Welbers:

```{r}
library(tokenbrowser)
meta = select(assignments, doc_id, year, president, party)
categorical_browser(assignments, meta=meta, category=as.factor(assignments$topic),  token_col="word") |>
  browseURL()
```