# RMarkdown parameters

RMarkdown has many features, as can be seen on this convenient [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf).
Here we'll discuss some usefull *codeblock parameters*.

Paremeters are given between the curly brackets above a codeblock, just after the r (see below for examples). There are many usefull parameters, but for now we'll focus on:

| parameter   | effect                                                         |
|-------------|----------------------------------------------------------------|
| eval = F    | don't evaluate (i.e. execute/run) the code, but do show output |
| echo = F    | don't show code, but do show output                            |
| message = F | don't show messages                                            |
| warning = F | don't show warnings                                            |  
| cache = T   | remember output, so it doensn't need to be evaluated the next  time. Resets when the code is changed |

### eval = F

The following code is not evaluated. This is for instance usefull to show that a certain package needs to be installed, without actually installing it. The following package does not exist, but the code is not evaluated so there is no error.

```{r, eval=F}
install.packages('not_even_a_real_package')
```

### echo = F

The code is 10*10, but only the output (100) is shown. This is usefull for actual reports, where the code is not important but the output is.
!!! NOTE: for the lab reports, do not use this, because I do want to see the code

```{r, echo=F}
10*10
```

### message = F

R operations can return messages (red text that is not a warning or error). This is usefull when working with code, but not pretty in reports. With message = F you can turn it off.

In the following 2 blocks, we'll first run the code normally, and the second time with message = F

```{r}
message("message example")
```

```{r, message=F}
message("You won't see me")
```

### warning = F

Like message = F, but for warnings.

```{r}
warning("warning!!")
```

```{r, message=F, warning=F}
warning("You won't see me")
```

### cache = T

If cache is TRUE, the output is stored, so that the next time you perform knitr, it does not have to be calculated/downloaded/etc. again. 
If you change the code (even a little bit) it forgets the results, and re-calculates on the next run.

In this example, the Sys.sleep() function let's R sleep for 10 seconds.
After knitting this RMarkdown file the first time, do it again to see that it now returns the output immediately.

```{r, cache=T}
Sys.sleep(10)


print('I fell asleep for 10 seconds')
```
